---
title: "NetFunrankR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NetFunrankR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 15,
  fig.height = 12,
  fig.align = "center",
  dpi = 144,
  out.width = "100%"
)
```

# NetFunrankR Wrapper Tutorial

This vignette demonstrates how to use the `NetFunrankR` package for network-proximity empowered functional enrichment test and comparative performance with `hypergeometric distribution` based over-representation test - with the `NetFunrankR_wrapper`.

## Overview

The `NetFunrankR_wrapper` function performs two types of pathway enrichment analyses: - **ORA**: Traditional over-representation analysis (hypergeometric test) - **TW**: Topology-aware network proximity analysis

Results include Reports (CSV files) with ranked pathways and comparative visualization (ggplot2 object and PDFs).

## Installation

Ensure required dependencies are installed:

    ```{r}
    # devtools::install_github("Akmazad/NetFunrankR")
    ```


## Basic Usage

### Load Package

```{r message=F, warning=F}
library(dplyr)
library(data.table)
library(ggplot2)
library(NetFunrankR)
outdir <- "./NetFunrankR_results/"
```

```{r theme-myvignette, include=FALSE}
library(ggplot2)

theme_myvignette <- function(base_size = 8, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(size = rel(1.4), face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = rel(1.1), hjust = 0.5),
      axis.title = element_text(size = rel(1.1)),
      axis.text = element_text(size = rel(1)),
      legend.title = element_text(size = rel(1.1)),
      legend.text = element_text(size = rel(1)),
      plot.margin = margin(10, 10, 10, 10),
      panel.grid.major = element_line(color = "gray80", size = 0.2),
      panel.grid.minor = element_blank()
    )
}

theme_set(theme_myvignette())
```

### Load example disease genes

The package comes with an example vector of disease genes (COPD) named example_enrich.df. Let's observe the head of this data..

```{r, message=F, warning=F}
print(head(COPD_proteinCodingGenes))
```

### Perform analysis

With the given disease genes, a wrapper function (`NetFunrankR_wrapper`) can be invoked to run the whole analysis, and the results (reports, plots - both pdf and ggplot2 objects) are saved in a given directory (default: `./NetFunrankR_Results/`).

```{r out.width="100%", out.height="100%", message=F, warning=F}
NetFunrankR_wrapper(
  disease_genes = COPD_proteinCodingGenes,
  PathwayRanker_Type = "Both",
  nPermute = 10,
  outdir = outdir
)
```

### Check results

#### Number of significantly enriched terms/pathways

This venn diagram shows counts of significantly enriched terms/pathways by the two methods

```{r out.width="100%", out.height="100%", message=F, warning=F}
load(paste0(outdir, "CommonFunctions_ORA_TW.vennplt.obj.RData"))
common.terms.vennplt.obj
```

#### Significantly enriched terms/pathways

This plot shows the terms/pathways that are significantly enriched (`pvalue < 0.05`) and commonly found between Network-Proximity based and Over-representation based methods

```{r out.width="100%", out.height="100%", message=F, warning=F}
load(paste0(outdir,"CommonFunctions_ORA_TW.barplt.obj.RData"))
common.terms.barplt.obj
```

#### Exclusive terms/pathways found by ORA

This plot shows the terms/pathways that are exclusively enriched (`pvalue < 0.05`) via ORA method

```{r out.width="100%", out.height="100%", message=F, warning=F}
load(paste0(outdir,"Only_ORA_terms.plt.obj.RData"))
only_ORA_terms.plt.obj
```

#### Exclusive terms/pathways found by TW

This plot shows the terms/pathways that are exclusively enriched (`pvalue < 0.05`) via TW method

```{r out.width="100%", out.height="100%", message=F, warning=F}
load(paste0(outdir,"Only_TW_terms.plt.obj.RData"))
only_TW_terms.plt.obj
```

### Further configuration of results

The enrichment results from both the methods can be further controlled based on filtering `p.adjust < 0.05` instead of the nominal `pvalue`.

```{r}
    rankedPathways_ORA <- fread(paste0(outdir,"KEGG_PE_ORA_result.csv"))
    rankedPathways_TW <- fread(paste0(outdir,"KEGG_PE_TopoAw_result.csv"))
    rankedPathways_ORA_filt <- rankedPathways_ORA %>% dplyr::filter(p.adjust < 0.05)
    rankedPathways_TW_filt <- rankedPathways_TW  %>% dplyr::filter(p.adjust < 0.05)
```

Then the results can be visualized accordingly.

```{r fig.width=12, fig.height=10, fig.align='center', message=F, warning=F}
library(ggvenn)
library(tidyr)
library(forcats)
library(ggplot2)
library(cowplot)

# venn diagram
p1 <- plot_venn(
        vec1 = rankedPathways_ORA_filt %>% dplyr::pull(Description),
        vec2 = rankedPathways_TW_filt %>% dplyr::pull(Description))
# double-bar plot
p2 <- inner_join(
        rankedPathways_ORA_filt,
        rankedPathways_TW_filt,
        by = "Description") %>%
  plot_double_bar()
# exclusive ORA 
p3 <- anti_join(
        rankedPathways_ORA_filt,
        rankedPathways_TW_filt,
        by = c("Description" = "Description")) %>%
  plot_lollipop(color_palette = c("#616530FF", "#dce58f"))
# exclusive TW
p4 <- anti_join(
        rankedPathways_TW_filt,
        rankedPathways_ORA_filt,
        by = c("Description" = "Description")) %>%
  plot_lollipop(color_palette = c("#CC8214FF", "#f6e8b1"))
p1_p2 <- plot_grid(p1,p2,
                            labels = c("a","b"),
                            rel_heights = c(5,6),
                            ncol = 1)
final_plot <- plot_grid(p1_p2,p3,p4,
                   # rel_heights = c(1,1),
                   rel_widths = c(1,1,1),
                   labels = c("", "c", "d"), 
                   ncol = 3,
                   align = "l"
                   )
# Display the combined plot
  ggsave(paste0(outdir,"all_plots_combined.pdf"), 
         plot = final_plot, 
         # device = cairo_pdf, 
         width = 20, height = 10, units = "in")
  final_plot
```

### Performance analysis on smaller input

With the given disease genes, a wrapper function (`NetFunrankR_wrapper`) can be invoked to run the whole analysis, and the results (reports, plots - both pdf and ggplot2 objects) are saved in a given directory (default: `./NetFunrankR_Results_small/`).

```{r out.width="100%", out.height="100%", message=F, warning=F}
outdir <- "./NetFunrankR_Results_small/"
NetFunrankR_wrapper(
  disease_genes = COPD_proteinCodingGenes %>% base::sample(size = 6, replace = FALSE),
  PathwayRanker_Type = "Both",
  nPermute = 10,
  outdir = outdir
)

```

We want to observe if the Network-proximity based enrichment test (TW) performs better than ORA when the input genesets are random and small-sized.

```{r}
    rankedPathways_ORA <- fread(paste0(outdir,"KEGG_PE_ORA_result.csv"))
    rankedPathways_TW <- fread(paste0(outdir,"KEGG_PE_TopoAw_result.csv"))
    rankedPathways_ORA_filt <- rankedPathways_ORA %>% dplyr::filter(Count > 0) %>% # this is important as sometimes functions with 0 overlap gets a nominal p-value
      dplyr::filter(pvalue < 0.05)
    rankedPathways_TW_filt <- rankedPathways_TW  %>% dplyr::filter(pvalue < 0.05)
```

Then the results can be visualized accordingly.

```{r fig.width=12, fig.height=10, fig.align='center', message=F, warning=F}
library(ggvenn)
library(tidyr)
library(forcats)
library(ggplot2)
library(cowplot)

# venn diagram
p1 <- plot_venn(
        vec1 = rankedPathways_ORA_filt %>% dplyr::pull(Description),
        vec2 = rankedPathways_TW_filt %>% dplyr::pull(Description))
# double-bar plot
p2 <- dplyr::inner_join(
        rankedPathways_ORA_filt,
        rankedPathways_TW_filt,
        by = "Description") %>%
  plot_double_bar()
# exclusive ORA 
p3 <- dplyr::anti_join(
        rankedPathways_ORA_filt,
        rankedPathways_TW_filt,
        by = c("Description" = "Description")) %>%
  plot_lollipop(color_palette = c("#616530FF", "#dce58f"))
# exclusive TW
p4 <- dplyr::anti_join(
        rankedPathways_TW_filt,
        rankedPathways_ORA_filt,
        by = c("Description" = "Description")) %>%
  plot_lollipop(color_palette = c("#CC8214FF", "#f6e8b1"))
p1_p2 <- cowplot::plot_grid(p1,p2,
                            labels = c("a","b"),
                            rel_heights = c(5,6),
                            ncol = 1)
final_plot <- cowplot::plot_grid(p1_p2,p3,p4,
                   # rel_heights = c(1,1),
                   rel_widths = c(1,1,1),
                   labels = c("", "c", "d"), 
                   ncol = 3,
                   align = "l"
                   )
# Display the combined plot
  ggsave(paste0(outdir,"all_plots_combined.pdf"), 
         plot = final_plot, 
         # device = cairo_pdf, 
         width = 20, height = 10, units = "in")
  final_plot
```

# Conclusion

In this vignette, we demonstrated the basic functionality of the NetFunrankR package. We loaded example disease genes, performed both Over-representation (ORA) and topology-aware (TW) network-proximity methods for functional term/pathway enrichment, compare, report and visualize results. For more detailed analysis, refer to the package documentation.

# References

-   Network proximity method: Menche et al., Science (2015)
-   STRING DB: <https://string-db.org>
-   ORA implementation: clusterProfiler (Yu et al., 2012)

------------------------------------------------------------------------

*For detailed methodology, see package documentation: `?NetFunrankR_wrapper`*
