---
title: "NetFunrankR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NetFunrankR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# NetFunrankR Wrapper Tutorial
This vignette demonstrates how to use the `NetFunrankR` package for network-proximity empowered functional enrichment test and comparative performance with `hypergeometric distribution` based over-representation test - with the `NetFunrankR_wrapper`.

## Overview
The `NetFunrankR_wrapper` function performs two types of pathway enrichment analyses:
 1. **ORA**: Traditional over-representation analysis (hypergeometric test)
 2. **TW**: Topology-aware network proximity analysis

Results include Reports (CSV files) with ranked pathways and comparative visualization (ggplot2 object and PDFs).

## Installation
Ensure required dependencies are installed:

# ```{r setup, message=FALSE, warning=F}
# # Install the package (if not public, use `devtools::install_local()`):
# devtools::install_github("Akmazad/NetFunrankR")
# ```

## Basic Usage

### Load Package 
```{r setup, message=F, warning=F}
library(dplyr)
library(data.table)
library(ggplot2)
library(NetFunrankR)
```

### Load example disease genes
The package comes with an example vector of disease genes (COPD) named example_enrich.df. Let's observe the head of this data..

```{r, message=F, warning=F}
print(head(COPD_proteinCodingGenes))
```


### Perform analysis
With the given disease genes, a wrapper function (`NetFunrankR_wrapper`) can be invoked to run the whole analysis, and the results (reports, plots - both pdf and ggplot2 objects) are saved in a given directory (default: `./NetFunrankR_Results/`).

```{r out.width="100%", out.height="100%", message=F, warning=F}
NetFunrankR_wrapper(
  disease_genes = COPD_proteinCodingGenes,
  PathwayRanker_Type = "Both",
  nPermute = 10,
  outdir = "./NetFunrankR_Results/"
)
```

### Check results
#### Number of significantly enriched terms/pathways
This venn diagram shows counts of significantly enriched terms/pathways by the two methods
```{r out.width="100%", out.height="100%", message=F, warning=F}
load("./NetFunrankR_results/CommonFunctions_ORA_TW.vennplt.obj.RData")
common.terms.vennplt.obj
```

#### Significantly enriched terms/pathways
This plot shows the terms/pathways that are significantly enriched (`pvalue < 0.05`) and commonly found between Network-Proximity based and Over-representation based methods
```{r out.width="100%", out.height="100%", message=F, warning=F}
load("./NetFunrankR_results/CommonFunctions_ORA_TW.barplt.obj.RData")
common.terms.barplt.obj
```

#### Exclusive terms/pathways found by ORA
This plot shows the terms/pathways that are exclusively enriched (`pvalue < 0.05`) via ORA method
```{r out.width="100%", out.height="100%", message=F, warning=F}
load("./NetFunrankR_results/Only_ORA_terms.plt.obj.RData")
only_ORA_terms.plt.obj
```

#### Exclusive terms/pathways found by TW
This plot shows the terms/pathways that are exclusively enriched (`pvalue < 0.05`) via TW method
```{r out.width="100%", out.height="100%", message=F, warning=F}
load("./NetFunrankR_results/Only_TW_terms.plt.obj.RData")
only_TW_terms.plt.obj
```
### Further configuration of results
The enrichment results from both the methods can be further controlled based on filtering `p.adjust < 0.05` instead of the nominal `pvalue`. 

```{r}
    rankedPathways_ORA <- fread("NetFunrankR_results/KEGG_PE_ORA_result.csv")
    rankedPathways_TW <- fread("NetFunrankR_results/KEGG_PE_TopoAw_result.csv")
    rankedPathways_ORA_filt <- rankedPathways_ORA %>% dplyr::filter(p.adjust < 0.05)
    rankedPathways_TW_filt <- rankedPathways_TW  %>% dplyr::filter(p.adjust < 0.05)
```

Then the results can be visualized accordingly.

```{r fig.width=12, fig.height=10, fig.align='center', message=F, warning=F}
library(ggvenn)
library(tidyr)
library(forcats)
library(ggplot2)
library(cowplot)

# venn diagram
p1 <- plot_venn(
        vec1 = rankedPathways_ORA_filt %>% dplyr::pull(Description),
        vec2 = rankedPathways_TW_filt %>% dplyr::pull(Description))
# double-bar plot
p2 <- dplyr::inner_join(
        rankedPathways_ORA_filt,
        rankedPathways_TW_filt,
        by = "Description") %>%
  plot_double_bar()
# exclusive ORA 
p3 <- dplyr::anti_join(
        rankedPathways_ORA_filt,
        rankedPathways_TW_filt,
        by = c("Description" = "Description")) %>%
  plot_lollipop(color_palette = c("#616530FF", "#dce58f"))
# exclusive TW
p4 <- dplyr::anti_join(
        rankedPathways_TW_filt,
        rankedPathways_ORA_filt,
        by = c("Description" = "Description")) %>%
  plot_lollipop(color_palette = c("#CC8214FF", "#f6e8b1"))
cowplot::plot_grid(p1,p2,p3,p4,
                   # rel_heights = c(1,1),
                   # rel_widths = c(1,1), 
                   labels = c("a", "b", "c", "d"), 
                   ncol = 2,
                   align = "l"
                   )
```

### Performance analysis on smaller input
With the given disease genes, a wrapper function (`NetFunrankR_wrapper`) can be invoked to run the whole analysis, and the results (reports, plots - both pdf and ggplot2 objects) are saved in a given directory (default: `./NetFunrankR_Results_small/`).

```{r out.width="100%", out.height="100%", message=F, warning=F}
NetFunrankR_wrapper(
  disease_genes = COPD_proteinCodingGenes %>% head(6),
  PathwayRanker_Type = "Both",
  nPermute = 10,
  outdir = "./NetFunrankR_Results_small/"
)
```


# Conclusion
In this vignette, we demonstrated the basic functionality of the NetFunrankR package. We loaded example disease genes, performed both Over-representation (ORA) and topology-aware (TW) network-proximity methods for functional term/pathway enrichment, compare, report and visualize results. For more detailed analysis, refer to the package documentation.

# References
 - Network proximity method: Menche et al., Science (2015)
 - STRING DB: https://string-db.org
 - ORA implementation: clusterProfiler (Yu et al., 2012)

 ---
 *For detailed methodology, see package documentation: `?NetFunrankR_wrapper`*




